Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-02 sections/eval.tex
Update on Overleaf.

\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
\begin{figure}
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    @finch begin
      C .= 0
      for j=_
        for i=_
          for k=_
            C[i, j] += AT[k, i] * B[k, j]
          end
        end
      end
      return C
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    w = Tensor(SparseByteMap(Element(0)))
    @finch begin
      C .= 0
      for j=_
        w .= 0
        for k=_
          for i=_
            w[i] += A[i, k] * B[k, j]
          end
        end
        for i=_
          C[i, j] = w[i]
        end
      end
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    w = Tensor(SparseHash(SparseHash(Element(0))))
    @finch begin
      w .= 0
      for k=_
        for j=_
          for i=_
            w[i, j] += A[i, k] * BT[j, k]
          end
        end
      end
      C .= 0
      for j=_, i=_
        C[i, j] = w[i, j]
      end
    end
    \end{minted}
    \end{minipage}
    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
    \caption{A comparison of several matrix multiplication algorithms between
    Finch and Taco. On left, we use smaller matrices, ordered from small to big
    dimension. Note that inner products necessarily requires $O(n^2)$ work and
    TACO's outer products format is dense. Finch can use a sparse outer products
    format and thus has an asymptotic advantage that becomes evident as the
    output dimensions grow. On right, we use only gustavson's algorithm and
    compare on larger matrices.}
    \label{fig:spgemm}
\end{figure}

@@

\begin{wrapfigure}{r}{0.41\linewidth}
    \begin{minipage}[t]{0.18\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _, i = _
              y[i] += A[i, j] * x[j]
            end
        \end{minted}
        \vspace{24pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _, i = _
              y[j] += A[i, j] * x[i]
            end
        \end{minted}
    \end{minipage}\hfill%
    \begin{minipage}[t]{0.22\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _
              let x_j = x[j]
                y_j .= 0
                for i = _
                  let A_ij = A[i, j]
                    y[i] += x_j * A_ij
                    y_j[] += A_ij * x[i]
                  end
                end
                #D is the diagonal
                y[j] += y_j[] + D[j] * x_j
              end
            end
        \end{minted}
    \end{minipage}
    \caption{Finch row-major, column-major and symmetric SpMV Programs}
    \label{spmv_programs}
\end{wrapfigure}

\begin{table}[htbp]
    \scriptsize
    \centering
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
        \hline
        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
        \cline{3-4} 
        & & Pattern & sparse, boolean-valued matrices \\
        \cline{2-4} 
        & SparseVBL & Element & real-valued matrices with blocked structure \\
        \cline{2-4}
        & SparseBand & Element & real-valued matrices with diagonal band \\
        \cline{2-4}
        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
        \cline{3-4} 
        & & Pattern & matrix with runs of true or false \\
        \hline 
    \end{tabular}
\end{table}

@@

\begin{table}[htbp]
    \centering
    \scriptsize
    \caption{Transposed SpMV Sample Matrices}
    \label{tab:transposed_spmv_sample_matrices}
    \begin{tabular}{|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|}
        \hline
         & \textbf{HB/saylr4} & \textbf{Norris/heart3} & \textbf{large\_band} & \textbf{SNAP/as-735} \\
        \hline
        \textbf{Spy} & \includegraphics[width=\linewidth]{spmv_matrices/saylr4.png} & \includegraphics[width=\linewidth]{spmv_matrices/heart3.png} & \includegraphics[width=\linewidth]{spmv_matrices/toeplitz_large_band.png} & \includegraphics[width=\linewidth]{spmv_matrices/as-735.png} \\
        \hline 
        \textbf{Group / Name} & HB/saylr4 & Norris/heart3 & large\_band & SNAP/as-735 \\
        \hline
        \textbf{Dimensions} & 3,564 x 3,564 (22,316) & 2,339 x 2,339 (680,341) & 10,000 x 10,000 (1,999,900) & 7,716 x 7,716 (26,467) \\
        \hline
        \textbf{Best Finch Format} & Symmetric SparseList & SparseVBL & SparseBand & Symmetric SparseList-Pattern \\
        \hline
    \end{tabular}
\end{table}

@@

\subsection{Implementing Numpy's Array API in Finch}
In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency. They generally either limit the dimensionality to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats. Further, existing work doesn't support the fusion of arbitrary operators which can have a drastic impact on performance as we show in Fig. (\kyle{PUT FIG HERE}). We believe that a flexible, online compiler like Finch which can manage the complexity of arbitrary operations between inputs with a wide variety of formats is the secret ingredient needed to make the array API performant for structured data.

There is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both the output and the algorithm. The naive approach to solving this would be to explicitly define a mapping from each function in the API to a Finch program which eagerly produces the output. This approach results in a simple translation procedure but misses crucial opportunities for optimization across a chain of function calls and needlessly materializes large intermediate results. To avoid this, we have implemented a lazy evaluation approach that collects API calls and then computes output when requested by the user. This is implemented through 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which lowers Finch Logic programs to one or more Finch programs while making heuristic decisions about output format, loop ordering, and protocols.

\subsubsection{Finch Logic}
The expression fragment of the logical language takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor. This means that every operator (with the exception of $\finchtable$ and $\finchalias$ which are analogous to the scan operator in RA) takes in an indexed tensor and outputs an indexed tensor. Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate the order. This ordering is necessary for two reasons 1) the outputs of a function in the Array API must match a defined order on the dimensions based on the order of the inputs' dimensions 2) to benefit from concordant iteration we must ensure that the operands to a mapjoin have compatible index orders.

To express materialization, reuse of common sub-expressions, and multiple outputs, we define queries and plans. A query is the assignment of the output of an expression to a name. We use this assignment to denote that we materialize an expression in memory. Later queries can then access the result of earlier queries through the alias operator which allows multiple queries to benefit from a shared computation. A plan is a sequence of queries and it produces a set of tensors as output.

\begin{align*}
\footnotesize
    \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
     \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
    \finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
     expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
\end{align*}

Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language. Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can access an index as a scalar to implement $argmin$. For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
\begin{align*}
\footnotesize
&\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
&\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
&\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
\end{align*}
Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops. In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels. However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.

@@

\begin{figure}
\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
\end{tabular}
\vspace{-12pt}
\caption{Performance of Finch Logic for common kernels.}
\end{figure}

>>>>

\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
\begin{figure}
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    @finch begin
      C .= 0
      for j=_
        for i=_
          for k=_
            C[i, j] += AT[k, i] * B[k, j]
          end
        end
      end
      return C
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    w = Tensor(SparseByteMap(Element(0)))
    @finch begin
      C .= 0
      for j=_
        w .= 0
        for k=_
          for i=_
            w[i] += A[i, k] * B[k, j]
          end
        end
        for i=_
          C[i, j] = w[i]
        end
      end
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    w = Tensor(SparseHash(SparseHash(Element(0))))
    @finch begin
      w .= 0
      for k=_
        for j=_
          for i=_
            w[i, j] += A[i, k] * BT[j, k]
          end
        end
      end
      C .= 0
      for j=_, i=_
        C[i, j] = w[i, j]
      end
    end
    \end{minted}
    \end{minipage}
    \vspace{-12pt}
    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
    \vspace{-12pt}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
    \vspace{-8pt}
    \caption{A comparison of several matrix multiplication algorithms between
    Finch and Taco. On left, we use smaller matrices, ordered from small to big
    dimension. Note that inner products necessarily requires $O(n^2)$ work and
    TACO's outer products format is dense. Finch can use a sparse outer products
    format and thus has an asymptotic advantage that becomes evident as the
    output dimensions grow. On right, we use only gustavson's algorithm and
    compare on larger matrices.}
    \label{fig:spgemm}
    \vspace{-12pt}
\end{figure}

@@

\begin{wrapfigure}{r}{0.41\linewidth}
    \begin{minipage}[t]{0.18\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _, i = _
              y[i] += A[i, j] * x[j]
            end
        \end{minted}
        \vspace{24pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _, i = _
              y[j] += A[i, j] * x[i]
            end
        \end{minted}
    \end{minipage}\hfill%
    \begin{minipage}[t]{0.22\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            y .= 0
            for j = _
              let x_j = x[j]
                y_j .= 0
                for i = _
                  let A_ij = A[i, j]
                    y[i] += x_j * A_ij
                    y_j[] += A_ij * x[i]
                  end
                end
                #D is the diagonal
                y[j] += y_j[] + D[j] * x_j
              end
            end
        \end{minted}
    \end{minipage}
    \vspace{-8pt}
    \caption{Finch row-major, column-major and symmetric SpMV Programs}
    \label{spmv_programs}
    \vspace{-12pt}
\end{wrapfigure}

\begin{table}[htbp]
    \scriptsize
    \centering
    \vspace{-12pt}
    \caption{SpMV Tensor Formats}
    \vspace{-12pt}
    \label{spmv_tensor_formats}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
        \hline
        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
        \cline{3-4} 
        & & Pattern & sparse, boolean-valued matrices \\
        \cline{2-4} 
        & SparseVBL & Element & real-valued matrices with blocked structure \\
        \cline{2-4}
        & SparseBand & Element & real-valued matrices with diagonal band \\
        \cline{2-4}
        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
        \cline{3-4} 
        & & Pattern & matrix with runs of true or false \\
        \hline 
    \end{tabular}
    \vspace{-8pt}
\end{table}

@@

\begin{table}[htbp]
    \centering
    \scriptsize
    \caption{Transposed SpMV Sample Matrices}
    \vspace{-12pt}
    \label{tab:transposed_spmv_sample_matrices}
    \begin{tabular}{|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|}
        \hline
         & \textbf{HB/saylr4} & \textbf{Norris/heart3} & \textbf{large\_band} & \textbf{SNAP/as-735} \\
        \hline
        \textbf{Spy} & \includegraphics[width=\linewidth]{spmv_matrices/saylr4.png} & \includegraphics[width=\linewidth]{spmv_matrices/heart3.png} & \includegraphics[width=\linewidth]{spmv_matrices/toeplitz_large_band.png} & \includegraphics[width=\linewidth]{spmv_matrices/as-735.png} \\
        \hline 
        \textbf{Group / Name} & HB/saylr4 & Norris/heart3 & large\_band & SNAP/as-735 \\
        \hline
        \textbf{Dimensions} & 3,564 x 3,564 (22,316) & 2,339 x 2,339 (680,341) & 10,000 x 10,000 (1,999,900) & 7,716 x 7,716 (26,467) \\
        \hline
        \textbf{Best Finch Format} & Symmetric SparseList & SparseVBL & SparseBand & Symmetric SparseList-Pattern \\
        \hline
    \end{tabular}
    \vspace{-8pt}
\end{table}

@@

\subsection{Implementing Numpy's Array API in Finch}
In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency. They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats. Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Fig. \ref{fig:fusion}. We believe that a flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the secret ingredient needed to make the array API performant for structured data. To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language. This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.

\subsubsection{Finch Logic}
The expression fragment of the logical language takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor. This means that operators take in a set of indexed tensors and output an indexed tensor. Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order. To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$. A $\finchquery$ assigns the output of an expression to a name. We use this to denote that we materialize an expression. Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation. A plan is a sequence of queries and outputs a set of tensors.

\begin{align*}
    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
\end{align*}

Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language. Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can access an index as a scalar to implement $argmin$. For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
\begin{align*}
\footnotesize &\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
\footnotesize &\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
\footnotesize &\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
\end{align*}
Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops. In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels. However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.

@@

\begin{figure}
\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
\end{tabular}
\vspace{-12pt}
\label{fig:fusion}
\caption{Performance of Finch Logic for common kernels.}
\end{figure}
